---
title: "Cancer Incidence Geographic Distribution Analysis"
author: "Olisaemeka Charles Okolo"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This report analyzes state-level cancer incidence rates across the United States
using publicly available CDC WONDER registry data (2018–2022). All rates are
age-adjusted per 100,000 population using the 2000 U.S. Standard Million,
enabling fair cross-state comparisons independent of age distribution differences.

## Load and Clean Data

```{r load-data}
library(tidyverse)
library(maps)
library(viridis)
library(scales)
library(knitr)

raw <- read_tsv(
  "US Puerto Rico Cancer Statistics 1999-2022.tsv",
  col_types = cols(.default = "c"),
  na = c("", "NA", "Suppressed", "Missing", "Not Applicable")
)

cancer <- raw %>%
  filter(!is.na(States), States != "Total") %>%
  select(
    state      = States,
    count      = Count,
    population = Population,
    rate       = `Age-Adjusted Rate`
  ) %>%
  mutate(
    rate       = as.numeric(rate),
    count      = as.numeric(gsub(",", "", count)),
    population = as.numeric(gsub(",", "", population)),
    state_lower = tolower(state)
  ) %>%
  filter(!is.na(rate))

kable(
  head(cancer %>% select(state, count, population, rate), 10),
  caption = "Sample: cleaned CDC cancer incidence data (2018–2022)",
  col.names = c("State", "Case count", "Population", "Age-adjusted rate")
)
```

## Geographic Distribution

```{r choropleth, fig.width=10, fig.height=6}
us_states <- map_data("state")

map_joined <- us_states %>%
  left_join(cancer, by = c("region" = "state_lower"))

ggplot(map_joined, aes(x = long, y = lat, group = group, fill = rate)) +
  geom_polygon(color = "white", linewidth = 0.3) +
  scale_fill_viridis(
    option = "plasma",
    name   = "Age-adjusted rate\nper 100,000",
    labels = comma
  ) +
  coord_fixed(1.3) +
  theme_void() +
  theme(
    plot.title    = element_text(size = 16, face = "bold", margin = margin(b = 8)),
    plot.subtitle = element_text(size = 12, color = "gray40", margin = margin(b = 16)),
    legend.position = "right"
  ) +
  labs(
    title    = "Cancer Incidence Rates by State",
    subtitle = "Age-adjusted rates per 100,000 population, 2018–2022 · CDC WONDER"
  )
```

## Top and Bottom 10 States

```{r bar-charts, fig.width=11, fig.height=5}
library(gridExtra)

top10 <- cancer %>% slice_max(rate, n = 10) %>% mutate(state = fct_reorder(state, rate))
bot10 <- cancer %>% slice_min(rate, n = 10) %>% mutate(state = fct_reorder(state, rate))

p1 <- ggplot(top10, aes(x = state, y = rate, fill = rate)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rate), hjust = -0.2, size = 3, color = "gray30") +
  scale_fill_viridis(option = "plasma") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  theme_minimal() +
  labs(title = "10 highest incidence states", x = NULL, y = "Age-adjusted rate per 100,000")

p2 <- ggplot(bot10, aes(x = state, y = rate, fill = rate)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rate), hjust = -0.2, size = 3, color = "gray30") +
  scale_fill_viridis(option = "plasma") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  theme_minimal() +
  labs(title = "10 lowest incidence states", x = NULL, y = "Age-adjusted rate per 100,000")

grid.arrange(p1, p2, ncol = 2)
```

## Summary Statistics

```{r summary}
cancer %>%
  summarise(
    `Mean rate`       = round(mean(rate, na.rm = TRUE), 1),
    `Median rate`     = round(median(rate, na.rm = TRUE), 1),
    `Highest rate`    = round(max(rate, na.rm = TRUE), 1),
    `Lowest rate`     = round(min(rate, na.rm = TRUE), 1),
    `Std deviation`   = round(sd(rate, na.rm = TRUE), 1),
    `States included` = n()
  ) %>%
  kable(caption = "National summary — age-adjusted cancer incidence rates per 100,000 (2018–2022)")
```

## Key Findings

```{r findings, results='asis'}
highest_state <- cancer %>% slice_max(rate, n = 1) %>% pull(state)
lowest_state  <- cancer %>% slice_min(rate, n = 1) %>% pull(state)
highest_rate  <- cancer %>% slice_max(rate, n = 1) %>% pull(rate)
lowest_rate   <- cancer %>% slice_min(rate, n = 1) %>% pull(rate)
gap           <- round(highest_rate - lowest_rate, 1)
nat_mean      <- round(mean(cancer$rate), 1)

cat(glue::glue("
- **{highest_state}** had the highest age-adjusted incidence rate at **{highest_rate}** per 100,000
- **{lowest_state}** had the lowest at **{lowest_rate}** per 100,000  
- The gap between highest and lowest states was **{gap}** per 100,000, highlighting significant geographic disparity in cancer burden
- The national age-adjusted mean was **{nat_mean}** per 100,000
"))
```

## Data Source

Centers for Disease Control and Prevention. *United States and Puerto Rico Cancer
Statistics, 1999–2022 Incidence*. CDC WONDER Online Database. U.S. Department of
Health and Human Services; 2025 release.
Retrieved from https://wonder.cdc.gov/cancer-v2022.html